% ── ABSTRACT ──────────────────────────────────────────────────────────────────
% Target length: ~300 words
% State: (1) the explanatory + operational framing,
%        (2) the Affinity Operator diagnostic in α-bar-language,
%        (3) the Coherence-Budget Allocation Theorem and benchmark,
%        (4) scope and hardware-contingency caveat.

The Relational Coherence framework (\RCF) does not alter the
predictions of quantum mechanics.
Its contribution is explanatory and operational: it recasts
decoherence, property definiteness, and tunable entanglement as
different aspects of a single coherence-budget process, in which
local relational potential is redistributed into environmental
correlation.
Concretely, I (i)~collect four standard identities of pointer-basis
coherence theory~\cite{Zurek2003,Baumgratz2014,Streltsov2017Review}
in the language of a normalized Hilbert--Schmidt coherence quantifier
$\bar\alpha(S,E) \in [0,1]$ grounded by einselection
(Bell's theorem~\cite{Bell1964} and its loophole-free experimental
verification~\cite{Aspect1982,Zeilinger2022Nobel} rule out local
non-contextual hidden-variable completions of quantum mechanics; I
adopt the relational reading of Rovelli~\cite{Rovelli1996,Rovelli2021}
as the conceptual setting, without claiming Bell's theorem singles it
out among the alternatives), and (ii)~derive an allocation theorem
for tunable Heisenberg-exchange gates~\cite{LossDiVincenzo1998} of
the form
\[
  \URA(\theta) \;=\; \exp\!\Bigl(\!-\tfrac{i\theta}{2}\,\bigl(
    \sigma_x{\otimes}\sigma_x + \sigma_y{\otimes}\sigma_y
    + \sigma_z{\otimes}\sigma_z\bigr)\!\Bigr),
\]
that turns the per-layer cost
$\bar\alpha^{\rm pre} - \bar\alpha^{\rm post}$ into a
compiler-relevant scheduling objective.

\textbf{Coherence-Budget Allocation Theorem.}
Under Markovian gate-time-proportional dephasing with effective
per-layer rates $\tilde\gamma_\ell$ and an entanglement target
$E_{\rm target} = \sum_\ell |\sin 2\theta_\ell| \in (0,L)$, the
unique cost-minimizing exchange-angle schedule is
\[
  \theta_\ell^{\!*} \;=\; \tfrac{1}{2}\arccos\!\Bigl(\!\min\!\Bigl(1,
    \;\tfrac{\tilde\gamma_\ell}{2\lambda^{\!*}}\Bigr)\!\Bigr),
\]
with shadow price $\lambda^{\!*}$ fixed by the budget.
The schedule routes more entanglement onto quieter layers and skips
layers whose $\tilde\gamma_\ell$ exceeds the global shadow price.

A reference scheduler implementing this theorem (\texttt{figures/rcb\_optimizer.py})
beats both fixed-gate and uniform-$\theta$ baselines in multi-layer
density-matrix simulation: against a concentrated full-entangler
baseline, the optimal schedule yields a final-state fidelity
advantage of up to $\Delta F = +0.040$ at $\gamma = 0.15$,
$E_{\rm target} = 1$, $L = 8$; against uniform-$\theta$ on
heterogeneous noise it yields $\Delta F = +0.025$ at the same target.
The framework's per-gate
predictions~\cite{Foxen2020,Abrams2020,Lacroix2020,Caldwell2018}
remain consistent with the experimental continuous-fSim advantage on
Sycamore-class hardware.
The advantage is contingent on direct exchange calibration; on
platforms where $\URA(\theta)$ must be decomposed into fixed native
primitives the advantage may shrink or vanish.
