% ── SECTION 1: INTRODUCTION (~500 words) ──────────────────────────────────────

\section{Introduction}
\label{sec:intro}

Bell's theorem~\cite{Bell1964,CHSH1969} together with its experimental
verification, culminating in the 2022 Nobel Prize in
Physics~\cite{Aspect1982,Zeilinger2022Nobel}, rules out local
non-contextual hidden-variable completions of quantum mechanics.
Several ontologically distinct programmes respond to this constraint:
Bohmian mechanics retains predetermined values at the cost of explicit
nonlocality~\cite{Bohm1980}; many-worlds dispenses with a single outcome
in favour of a universal wavefunction~\cite{Wallace2012}; QBism reads
the quantum state as an agent's degree of belief~\cite{FMS2014}; and
relational quantum mechanics (RQM)~\cite{Rovelli1996,Rovelli2021}
treats quantum properties as facts relative to the interacting systems.

The present paper works within the relational reading, without
claiming that Bell's theorem singles it out among the options above.
What we contribute is an explicit quantitative formulation of one
aspect of the relational picture~---~the dynamics of pointer-basis
coherence under decoherence~---~as a single computable scalar
quantity, the
\emph{scalar affinity} $\bar\alpha(S,E) \in [0,1]$, defined as the
normalized Hilbert--Schmidt coherence of the reduced state $\rho_S$
in the einselected pointer basis~\cite{Zurek2003,Schlosshauer2005}.
The Relational Coherence framework (\RCF) collects four standard
identities involving $\bar\alpha$, suitable as a per-layer
coherence-budget diagnostic in two-qubit circuits.
We stress at the outset that the \RCF~does not dispute the predictive
record of standard quantum mechanics; it offers a unified notation for
relating coherence, decoherence, and entanglement dynamics, and our
mathematical content reduces to standard pointer-basis coherence
theory~\cite{Baumgratz2014,Streltsov2017Review} grounded by einselection.

The \RCF~comprises four propositions: property indefiniteness (a
nonzero $\bar\alpha$ precludes a pointer-basis-diagonal $\rho_S$);
relational determination ($\bar\alpha$ decays exponentially under
Markovian dephasing in the pointer basis); coherence--affinity
correspondence ($\bar\alpha$ ranges continuously over $[0,1]$ with
extrema at full decoherence and the equal-weight pointer-basis
superposition); and coherence redistribution under unitary $S$--$E$
evolution (purity is conserved, mutual information grows, and local
off-diagonals are exported into environment-record orthogonality).
The proofs reduce to Cauchy--Schwarz, a Lindblad calculation, a
Schmidt-decomposition identity, and the Joos--Zeh / Zurek branching
argument~\cite{JoosZeh1985,Zurek2003}, in each case packaged so that
the role of $\bar\alpha$ as a circuit-level figure of merit is
explicit.

Continuously parameterized two-qubit exchange gates are a mature
hardware engineering line: from Loss--DiVincenzo
spin-qubit exchange~\cite{LossDiVincenzo1998}, through parametric
entangling gates on transmon platforms~\cite{Caldwell2018}, to the
continuous fSim/$XY$ families demonstrated on
Sycamore~\cite{Foxen2020,Abrams2020,Lacroix2020} and the
continuously parameterized $\mathcal{ZZ}$ (M\o{}lmer--S\o{}rensen)
gates studied for trapped-ion hardware~\cite{Yale2025NoiseAware}.
Noise-aware compilation exploiting continuous parameterization has
been demonstrated to improve circuit fidelity by routing heavier
entangling weight to better-performing gate pairs and removing
low-impact entangling operations~\cite{Yale2025NoiseAware}.
Building on this literature, we examine the isotropic Heisenberg-exchange
unitary
$\URA(\theta) = \exp[-i\theta(\sigma_x{\otimes}\sigma_x +
\sigma_y{\otimes}\sigma_y + \sigma_z{\otimes}\sigma_z)/2]$,
$\theta \in [0,\pi/2]$, which traces the main diagonal
$(c_1,c_2,c_3)=(\theta,\theta,\theta)$ of the Weyl chamber of
$\mathrm{SU}(4)/[\mathrm{SU}(2)\otimes\mathrm{SU}(2)]$~\cite{Zhang2003WeylChamber}
and attains maximum entanglement capability at $\theta = \pi/4$ (the
$\sqrt{\textsc{swap}}$ point).
We do not claim novelty for the unitary, which is the well-known
isotropic-exchange family; what the \RCF~contributes is a coherence-budget
framing for its parameter selection.
We emphasize that the gate parameter $\theta$ is a Hamiltonian evolution
angle, distinct from the scalar affinity $\bar\alpha \in [0,1]$; the
connection between the two is indirect (Sec.~\ref{sec:gate}), and the
\RCF~furnishes a conceptual warrant for a continuously tunable exchange
primitive, not a quantitative identity $\theta = \bar\alpha$.
We demonstrate via numerical simulation in
Qiskit~\cite{Qiskit2024} and Cirq~\cite{Cirq2025} that on hardware
with directly calibratable exchange coupling, $\URA(\theta < \pi/4)$
achieves measurable process-fidelity advantages over fixed-angle
entanglers in high-decoherence regimes, consistent with the
continuous-fSim experimental advantage reported on
Sycamore~\cite{Foxen2020}.

The technical contribution of this paper is the
\emph{Coherence-Budget Allocation Theorem} of
Sec.~\ref{sec:alloc}: under Markovian gate-time-proportional
decoherence with effective per-layer rates $\tilde\gamma_\ell$ and
total entangling target
$E_{\rm target} = \sum_\ell |\sin 2\theta_\ell|$, the unique
cost-minimizing exchange-angle schedule is
$\theta_\ell^\ast = \tfrac{1}{2}\arccos\bigl(\min(1,\tilde\gamma_\ell/(2\lambda^\ast))\bigr)$,
with shadow price $\lambda^\ast$ fixed by the budget.
The theorem yields three falsifiable predictions: distribute
entanglement across layers rather than concentrate it (uniform-noise
case); route entanglement onto quieter layers (heterogeneous case);
skip layers whose effective rate exceeds the global shadow price.
Qualitatively similar angle-routing and gate-removal behaviours have
been demonstrated empirically through compiler optimizations for
continuously parameterized $\mathcal{ZZ}$ gates on trapped-ion
hardware~\cite{Yale2025NoiseAware}; the present work differs by
deriving these behaviours analytically as the unique solution of a
convex coherence-budget program grounded in the einselection-derived
scalar affinity $\bar\alpha$ (Sec.~\ref{subsec:prior-compilation}).
A reference scheduler implementing the theorem
(\texttt{figures/rcb\_optimizer.py}) beats both fixed-gate and
uniform-$\theta$ baselines in multi-layer density-matrix simulation.

\subsection{From Substance Thinking to Relational-Affinity Thinking}
\label{sec:substance-to-relational}

The broader purpose of the \RCF~is not to replace standard quantum
mechanics, nor to deny the validity of Newtonian mechanics in its
classical domain.
Rather, it is to displace Newtonian \emph{substance}-thinking as the
default ontology carried into quantum theory.
In the Newtonian picture, one begins with separable bodies possessing
intrinsic properties, and relations are then introduced as secondary
interactions among already-defined things.
In the relational-affinity picture developed here, one begins instead
with interaction, coherence, and environment-relative definiteness;
what appear classically as separable objects are stable limiting
patterns produced by decoherence, einselection, and the suppression
of pointer-basis off-diagonals~\cite{Zurek2003,Schlosshauer2005}.

The scalar affinity $\bar\alpha(S,E)$ is introduced as a diagnostic
for this transition.
When $\bar\alpha > 0$, the reduced state $\rho_S$ retains pointer-basis
coherence and cannot be represented as a classical mixture of definite
pointer states (Proposition~\ref{prop:indefiniteness}).
When $\bar\alpha = 0$, $\rho_S$ is diagonal in the pointer basis
selected by the system--environment coupling, and the classical
object description becomes valid.
The \RCF~therefore does not treat classical separability as
fundamental; it treats separability as the $\bar\alpha \to 0$ limit
of relational decoherence (Proposition~\ref{prop:determination}).

This reframing is modest but important.
It does not claim that affinity is a new force, that quantum mechanics
requires a new dynamical law, or that relational quantum mechanics is
uniquely selected by Bell-type results.
Rather, it supplies an operational vocabulary for tracking how quantum
systems move from relational coherence to classical definiteness.
Newtonian mechanics remains an acceptable approximation for the
resulting stable patterns, but it is no longer the metaphysical
starting point.
The starting point is relation; the object is the emergent limit.

The paper is organized as follows.
Section~\ref{sec:background} reviews Bell's theorem, RQM, decoherence
and einselection, and the resource theory of coherence.
Section~\ref{sec:rct} states and proves the four \RCF~propositions.
Section~\ref{sec:gate} defines $\URA(\theta)$, locates it in the Weyl
chamber, and derives its entanglement capability.
Section~\ref{sec:sim} reports the per-gate numerical simulations.
Section~\ref{sec:alloc} introduces the Relational Coherence Cost,
proves the Coherence-Budget Allocation Theorem, and benchmarks the
resulting scheduler against fixed-gate and uniform-$\theta$ baselines.
Sections~\ref{sec:disc} and~\ref{sec:conc} discuss scope, limitations,
and conclude.
Full proofs appear in Appendix~\ref{app:rct-proof}.
