% ============================================================
%  Section II — Theoretical Background
%  "The Relational Coherence Framework"
%  Dr. Joshua Adams | 2026
% ============================================================

\section{\label{sec:background}Theoretical Background}

The Relational Coherence framework rests on two independent pillars of
modern quantum theory: the relational structure of quantum properties,
established theoretically by Bell's theorem~\cite{Bell1964} and
empirically by its experimental
tests~\cite{Aspect1982,Zeilinger2022Nobel} and articulated
philosophically by relational quantum
mechanics~\cite{Rovelli1996,Rovelli2021}, and the dynamical mechanism
of environmental decoherence~\cite{Zurek2003}, which governs how that
relational structure evolves toward the classical limit.
This section reviews each pillar in turn, concluding with Zurek's
theory of einselection~\cite{Zurek2003}, which supplies the physical
grounding for the pointer basis in which the Affinity Tensor is later
defined.

% ── II.A ─────────────────────────────────────────────────────────────────────

\subsection{\label{sec:bell}Bell Nonlocality and the Relational Structure
            of Quantum Properties}

The status of quantum properties --- whether they are intrinsic to
individual systems or constitutively relational --- was sharpened into
an empirical question by Bell's 1964 theorem~\cite{Bell1964}.
Einstein, Podolsky, and Rosen had argued in 1935 that the completeness
of quantum mechanics required the existence of elements of physical
reality corresponding to every observable whose value could be predicted
with certainty without disturbing the system~\cite{Einstein1935EPR}.
If such elements exist and are local --- that is, if the properties of a
system are determined independently of measurements performed at
space-like separation --- then any statistical predictions must satisfy
Bell's inequality.

The refined operational form of this argument, due to Clauser, Horne,
Shimony, and Holt~\cite{CHSH1969}, was tested experimentally by
Aspect, Grangier, and Roger in 1982~\cite{Aspect1982}, who reported
$S_{\rm exp} = 2.697 \pm 0.015$ in polarization-entangled photon pairs,
violating the CHSH bound $\lvert S\rvert \leq 2$ by more than
forty standard deviations.
Subsequent experiments across multiple platforms closed successive
loopholes --- fair sampling, locality, and freedom of choice --- in a
sequence of loophole-free tests using NV-center electron
spins~\cite{Hensen2015} and entangled
photons~\cite{Giustina2015,Shalm2015}, with the freedom-of-choice
loophole subsequently addressed by deriving measurement settings from
high-redshift quasar emission~\cite{Rauch2018}.
In 2022 the Nobel Prize in Physics recognized this body of work, with
Zeilinger's Nobel Lecture~\cite{Zeilinger2022Nobel} providing a
comprehensive account of loophole-free Bell tests.

The physical content of these results is a theorem-level constraint:
no theory of local, non-contextual hidden variables can reproduce the
quantum-mechanical predictions.
Viable ontologies must respond to this constraint.
As noted in Sec.~\ref{sec:intro}, four programmes do so~---~Bohmian
mechanics~\cite{Bohm1980}, many-worlds~\cite{Wallace2012},
QBism~\cite{FMS2014}, and relational quantum mechanics.
The present work adopts the relational reading because it provides
the most direct ontological correlate for the pointer-basis coherence
quantity that organizes our formalism, not because Bell's theorem
selects it among the alternatives above.

% ── II.B ─────────────────────────────────────────────────────────────────────

\subsection{\label{sec:rqm}Relational Quantum Mechanics}

The most direct ontological response to Bell nonlocality is relational
quantum mechanics (RQM), first developed systematically by
Rovelli~\cite{Rovelli1996} and later presented accessibly in
\textit{Helgoland}~\cite{Rovelli2021}.
RQM accepts the Bell result at face value and elevates it to a founding
principle: a quantum system $S$ does not possess definite physical
properties absolutely; it possesses properties only \emph{relative to}
another system $O$ with which it has interacted.
Distinct observers may assign different, mutually consistent quantum
states to the same system without contradiction, because those states are
relational facts, not absolute ones.

In Rovelli's notation, a quantum state is indexed to the observing
system: states are facts relative to interactions, not absolute
properties of $S$ alone.
For the mathematical work in this paper we adopt the standard joint
and reduced density-operator notation $\rho_{SE}$ and
$\rho_S = \mathrm{Tr}_E[\rho_{SE}]$ (Sec.~\ref{sec:rct},
Definition~\ref{def:rho}); the relational indexing enters not through
a separate notational object but through the einselected pointer basis
in which the Affinity Tensor is defined.
The relational network $\Gamma_S$ encodes the set of systems
with which $S$ has exchanged interactions, and it is this network that
determines which properties of $S$ are actualized and for whom.
In the present work $\Gamma_S$ is treated as a collection of
interaction events rather than a full graph-theoretic structure: the
Affinity Tensor and its scalar summary defined in
Sec.~\ref{sec:rct} depend on the joint state $\rho_{SE}$ at the time
of evaluation together with the pointer basis fixed by the
$S$-$E$ Hamiltonian, so the temporal ordering of prior interaction
events does not enter the present formalism.
The extension to a weighted interaction graph, in which edge weights
carry coupling strengths and event order may matter, is reserved for
future development.
Two consequences follow immediately.
First, the quantum state is not a representation of objective reality
independent of an observational context; it is an encoding of the
correlations between $S$ and $O$ that $O$ could in principle retrieve.
Second, entanglement is not a peculiarity to be explained away but the
\emph{canonical} form of relational existence: a system whose properties
are maximally correlated with another is, in the fullest quantum sense,
maximally related to it.
What is colloquially called ``spooky action at a distance'' is, on this
view, not a mysterious nonlocal connection between otherwise independent
objects; it is the direct signature of a constitutive relation ---
the two systems are not separately real things that happen to be
correlated, but are jointly and mutually defined through their
interaction history encoded in $\Gamma_S$.

This ontological picture leaves open a critical question: \emph{how}
does the web of relational facts stabilize into the appearance of a
shared classical world?
Rovelli's framework describes the logical structure of relational
facts but does not, by itself, supply the dynamical mechanism.
That mechanism is environmental decoherence.
Relational quantum mechanics tells us what quantum properties
\emph{are} --- facts relative to an observer --- while decoherence
theory tells us how they \emph{stabilize} into the classical-seeming
world that multiple observers agree upon.
To realize this mathematical structure, we must examine the primary
mechanism through which relational coherence is lost: environmental
decoherence.

% ── II.C ─────────────────────────────────────────────────────────────────────

\subsection{\label{sec:decoherence}Environmental Decoherence and Einselection}

A quantum system $S$ is never truly isolated.
Its Hilbert space $\mathcal{H}_S$ is perpetually coupled to an
environmental Hilbert space $\mathcal{H}_E$ of enormous dimension, and
this coupling drives the dynamical process of decoherence.
The evolution of the composite state $\rho_{SE}$ under a total
Hamiltonian $H = H_S + H_E + H_{\mathrm{int}}$ is, after tracing over
the environmental degrees of freedom, well approximated by the
Lindblad master equation~\cite{Lindblad1976,GKS1976,NielsenChuang2000}:

\begin{equation}
\label{eq:lindblad}
\frac{d\rho_S}{dt}
  = -\frac{i}{\hbar}\bigl[H_S,\, \rho_S\bigr]
    + \sum_k \!\left(
        L_k \rho_S L_k^{\dagger}
        - \tfrac{1}{2}\bigl\{ L_k^{\dagger} L_k,\, \rho_S \bigr\}
      \right),
\end{equation}

\noindent
where $\{L_k\}$ are the Lindblad jump operators encoding the
system-environment coupling and $\rho_S = \mathrm{Tr}_E[\rho_{SE}]$
is the reduced density matrix of $S$.
Equation~\eqref{eq:lindblad} is the most general Markovian,
trace-preserving, completely positive evolution for an open quantum
system.
The Markovian approximation --- valid when environmental correlation
times are short compared to the system's coherence time --- is standard
for superconducting qubit platforms operating at dilution-refrigerator
temperatures~\cite{Krantz2019}, and is adopted throughout this work.
Non-Markovian corrections introduce memory kernels that would complicate
the Affinity Tensor dynamics without qualitatively altering the
coherence-loss mechanism the \RCF~characterizes; their treatment is
deferred to future work.

The signature of decoherence is the exponential suppression of the
off-diagonal elements of $\rho_S$ in a preferred basis.
For a pure-dephasing environment, in which the jump operators $\{L_k\}$
of Eq.~\eqref{eq:lindblad} are diagonal in the pointer basis selected
by einselection (below), the dissipative term of the Lindblad equation
reduces to a simple decay of off-diagonal density-matrix elements while
the diagonal populations are preserved
(the explicit calculation appears in the proof of
Proposition~\ref{prop:determination}, Sec.~\ref{sec:prop2}).
For an initially pure state $\rho_S(0) = \vert\psi\rangle\langle\psi\vert$,
the coherences then decay as

\begin{equation}
\label{eq:decoherence-decay}
\langle i \vert \rho_S(t) \vert j \rangle
  \;=\;
  \langle i \vert \rho_S(0) \vert j \rangle\;
  e^{-\Gamma_{ij}\, t},
  \qquad i \neq j,
\end{equation}

\noindent
where the dephasing rate $\Gamma_{ij} > 0$ depends on coupling
strength and environmental correlation time.
On timescales $t \gg \Gamma_{ij}^{-1}$, all off-diagonal elements
vanish and $\rho_S$ becomes diagonal --- a classical probability
distribution over the basis states $\{\vert i\rangle\}$.
This is the quantum-to-classical transition, observable in mesoscopic
systems and increasingly relevant at the scale of near-term quantum
processors operating under non-negligible noise.

\subsubsection*{Einselection and the Pointer Basis}

A critical question remains: in which basis do the coherences of
Eq.~\eqref{eq:decoherence-decay} decay?
A unitary basis change can rotate any decohering basis into another, so
without a physical mechanism that singles out a preferred basis the
off-diagonal-decay statement is ambiguous.
Zurek's theory of \emph{einselection} (environment-induced
superselection)~\cite{Zurek2003} answers this from first principles
and is essential to grounding the Affinity Tensor defined in
Sec.~\ref{sec:rct}.

For the dephasing model simulated in Sec.~\ref{sec:sim}, the
system-environment coupling takes the prototypical form
$H_{\mathrm{int}} = g\,\sigma_z \otimes \hat{B}_E$,
where $\sigma_z$ is the Pauli-$Z$ operator on $S$ and
$\hat{B}_E$ is a bath observable; the two-qubit extension is
$H_{\mathrm{int}} = g\,(\sigma_z \otimes \sigma_z) \otimes \hat{B}_E$.
This diagonal coupling drives pure dephasing without energy relaxation
and directly determines the pointer basis via einselection.

The interaction Hamiltonian $H_{\mathrm{int}}$ generically commutes
with a preferred set of system observables whose eigenstates are
the \emph{pointer states} $\{\vert \pi_k \rangle\}$:

\begin{equation}
\label{eq:pointer}
\bigl[ H_{\mathrm{int}},\;
       \vert \pi_k \rangle\langle \pi_k \vert \bigr]
\;\approx\; 0.
\end{equation}

The commutation here is approximate: for finite-bath, finite-time, and
non-Markovian dynamics the pointer basis is not uniquely defined and
is selected through a predictability sieve in which residual
non-pointer-basis off-diagonals decay
faster~\cite{Schlosshauer2005,Zurek2003}.
We work in the idealization in which einselection has converged on a
sharp pointer basis; the basis-dependence of the Affinity Tensor below
(Sec.~\ref{sec:rct}) is then grounded by this idealization, and a more
careful treatment for non-Markovian or multi-bath situations is left
to future work.
Pointer states are dynamically stable under environmental
monitoring: they leave a robust imprint on $E$ without being
disturbed in return.
All superpositions of pointer states decohere into mixtures of pointer
states at the rates $\Gamma_{ij}$; the pointer states themselves
persist.
The environment thus \emph{selects} this basis --- hence
``einselection'' --- independently of any experimental choice.
This is a physically decisive point: the role traditionally assigned
to the ``observer'' is played here by the environment itself --- a
concrete physical system governed by the same Hamiltonian dynamics as
$S$, not an epistemological agent standing outside the formalism.
The classical world is not revealed by conscious observation; it is
\emph{produced} by physical interaction, in direct consonance with the
relational ontology of RQM.

Einselection has two consequences essential to the present work.
First, it provides a physically grounded, basis-independent definition
of classicality: a state is classical relative to environment $E$
precisely when it is a mixture of pointer states of $H_{\mathrm{int}}$.
Second, it specifies the natural basis for the Affinity Tensor
$\alpha(S, E)$: rather than an arbitrary computational basis,
we define $\alpha(S, E)$ in the pointer basis singled out by the
$S$-$E$ coupling (Sec.~\ref{sec:rct}).
This choice transforms the framework from a formal statement about
off-diagonal matrix elements into a description of how the classical
world emerges from the quantum one.

The full trajectory is therefore the following.
A system $S$ begins in a superposition of pointer states with its
off-diagonal density-matrix elements at their maximum value
($\bar{\alpha} = 1$); the $S$-$E$ interaction entangles them and
drives those coherences toward zero at rate $\Gamma_{ij}$; and in
the long-time limit the off-diagonal elements vanish entirely in the
pointer basis ($\bar{\alpha} = 0$), leaving a classical mixture of
pointer-state outcomes~\cite{Zurek2003}.
We distinguish two quantities along this trajectory, and the
distinction is essential to what follows.
The scalar affinity $\bar{\alpha}$ measures the \emph{internal
relational potential} of $S$~---~the magnitude of coherent superposition
locally available in $\rho_S$ for relational determination~---~whereas
the $S$-$E$ entanglement measures the \emph{actualized relational
connection} already encoded in the joint state $\rho_{SE}$.
At $\bar{\alpha} = 1$ the potential is maximal and the connection
to $E$ is zero; as decoherence proceeds the potential is expended into
$S$-$E$ correlations, and at $\bar{\alpha} = 0$ the potential is fully
spent and the connection is maximal.
The two endpoints of this trajectory, for an $S$ initialized in the
maximally coherent equal-weight pointer-basis superposition, are
%
\begin{equation}
\label{eq:complementarity}
  \bar{\alpha}\!:\;1 \;\longrightarrow\; 0,
  \qquad
  S(\rho_S)\!:\;0 \;\longrightarrow\; \log d,
\end{equation}
%
where $S(\rho_S) \equiv -\mathrm{Tr}[\rho_S \log \rho_S]$ is the
von Neumann entropy of the reduced state and $d = \dim \mathcal{H}_S$;
under the unitary $S{+}E$ evolution of Sec.~\ref{sec:rct}, $S(\rho_S)$
coincides with the bipartite entanglement entropy between $S$ and $E$
(Proposition~\ref{prop:conservation}, part~(ii)).
The scalar affinity $\bar{\alpha}$ is therefore the figure of merit
for tracking how much local pointer-basis coherence remains in
$\rho_S$ at any moment in a circuit's execution, distinct from
quantities such as concurrence or $S$-$E$ entanglement entropy that
characterize the joint state.
The Affinity Tensor tracks exactly this trajectory, and the
Relational Coherence framework formalizes it.
