% ── SECTION 4: THE RELATIONAL AFFINITY GATE (~1,000 words) ──────────────────

\section{The Relational Affinity Gate}
\label{sec:gate}

\subsection{Definition}

The \RCT~establishes that relational affinity~$\bar\alpha$ is the fundamental
dynamical quantity governing coherence and entanglement.
A gate that implements controlled entanglement should therefore be parameterized
directly by~$\bar\alpha$, rather than operating at a fixed entanglement
capability as conventional two-qubit gates do.

\begin{definition}[Relational Affinity Gate]
The \emph{Relational Affinity Gate} is the two-qubit unitary
\begin{equation}
  \URA(\alpha) = \exp\!\left(-\frac{i\alpha}{2}
    \bigl(\sigma_x \otimes \sigma_x
         + \sigma_y \otimes \sigma_y
         + \sigma_z \otimes \sigma_z\bigr)\right),
  \label{eq:URA}
\end{equation}
where $\alpha \in [0, \pi/2]$ is the Affinity Tensor parameter and
$\sigma_{x,y,z}$ are the Pauli matrices acting on $\mathbb{C}^2$.
\end{definition}

The generator $\hat{H} = \tfrac{1}{2}(\sigma_x\otimes\sigma_x +
\sigma_y\otimes\sigma_y + \sigma_z\otimes\sigma_z)$ is the isotropic
Heisenberg exchange interaction, which generates entanglement symmetrically
across all three Pauli axes.
The parameter~$\alpha$ plays the role of an interaction time or coupling
strength, and is identified with the scalar Affinity Tensor value
$\bar\alpha$ of the two-qubit system.

In the computational basis $\{\ket{00}, \ket{01}, \ket{10}, \ket{11}\}$,
the gate takes the explicit form
\begin{equation}
  \URA(\alpha) =
  \begin{pmatrix}
    e^{-i\alpha/2} & 0 & 0 & 0 \\
    0 & e^{i\alpha/2}\cos\alpha & {-i}\,e^{i\alpha/2}\sin\alpha & 0 \\
    0 & {-i}\,e^{i\alpha/2}\sin\alpha & e^{i\alpha/2}\cos\alpha & 0 \\
    0 & 0 & 0 & e^{-i\alpha/2}
  \end{pmatrix}.
  \label{eq:URA-matrix}
\end{equation}
This form follows from the spectral decomposition of the generator $G =
\sigma_x\!\otimes\!\sigma_x + \sigma_y\!\otimes\!\sigma_y + \sigma_z\!\otimes\!\sigma_z$,
whose eigenvalues are $+1$ (triplet, three-fold: $\ket{00}$, $\ket{11}$,
$\ket{\Psi^+}$) and $-3$ (singlet: $\ket{\Psi^-}$).
Accordingly, $\URA(\alpha)$ acts on the two invariant subspaces independently:
the triplet states $\ket{00}$ and $\ket{11}$ each acquire the phase $e^{-i\alpha/2}$,
while states in the $\{\ket{01},\ket{10}\}$ subspace receive a mixture of
the triplet phase $e^{-i\alpha/2}$ and singlet phase $e^{3i\alpha/2}$,
producing the coherent rotation seen in the central $2\times2$ block.
At $\alpha = 0$, $\URA = \mathbf{I}_4$ (identity).
At $\alpha = \pi/4$, the gate achieves maximum entanglement capability
(see Section~\ref{subsec:weyl}).
At $\alpha = \pi/2$, $\URA$ reduces to the \textsc{swap} gate up to
local operations.

\subsection{Bell State Production}
\label{subsec:bell}

The gate's relational character is most directly illustrated by its
action on the computational basis state $\ket{01}$ at $\alpha = \pi/4$.
Because $\ket{01} = \tfrac{1}{\sqrt{2}}(\ket{\Psi^+} + \ket{\Psi^-})$
decomposes into equal triplet ($\ket{\Psi^+}$, eigenvalue $+1$) and
singlet ($\ket{\Psi^-}$, eigenvalue $-3$) components of the generator $G$,
the gate applies distinct phase factors to each branch.
Reading from Eq.~\eqref{eq:URA-matrix}:
\begin{equation}
  \URA(\pi/4)\ket{01}
    = \frac{e^{i\pi/8}}{\sqrt{2}}\bigl(\ket{01} - i\ket{10}\bigr),
  \label{eq:bell-production}
\end{equation}
a maximally entangled state with concurrence $\mathcal{C} = 1$,
consistent with the entanglement capability peak at $\alpha = \pi/4$
(Sec.~\ref{subsec:entcap}).
The global phase $e^{i\pi/8}$ is physically unobservable.

States lying entirely within the triplet sector~---~$\ket{00}$, $\ket{11}$,
and $\ket{\Psi^+}$~---~each acquire only the phase $e^{-i\alpha/2}$ under
$\URA(\alpha)$ and remain unentangled.
Entanglement generation therefore requires an input state with nonzero
singlet weight $\langle\Psi^-\vert\psi_{\rm in}\rangle \neq 0$.
All four Bell states are accessible from the computational basis via
$\URA(\pi/4)$ composed with appropriate single-qubit Pauli corrections,
confirming that $\URA$ spans the full entanglement capability of the
two-qubit Hilbert space.

\subsection{Weyl Chamber Analysis}
\label{subsec:weyl}

Two-qubit gates are classified up to local operations by their position in
the Weyl chamber of $\mathrm{SU}(4)/[\mathrm{SU}(2)\otimes\mathrm{SU}(2)]$,
parameterized by coordinates $(c_1, c_2, c_3) \in [0,\pi/2]^3$ with
$c_1 \geq c_2 \geq c_3 \geq 0$~\cite{Zhang2003WeylChamber}.
These coordinates are extracted from the \emph{local invariants} of the
two-qubit gate matrix.
We adopt the convention in which the generator is $\sum_k \sigma_k\!\otimes\!\sigma_k$
(rather than $\tfrac{1}{2}\sum_k \sigma_k\!\otimes\!\sigma_k$ as in some
references~\cite{Zhang2003WeylChamber}), so our coordinates are a factor of two
larger; in this convention \textsc{swap} sits at $(\pi/2, \pi/2, \pi/2)$.

For $\URA(\alpha)$, the Weyl chamber coordinates are:
\begin{equation}
  c_1(\alpha) = \alpha, \qquad
  c_2(\alpha) = \alpha, \qquad
  c_3(\alpha) = \alpha.
  \label{eq:weyl-coords}
\end{equation}
The trajectory $(c_1, c_2, c_3) = (\alpha, \alpha, \alpha)$ lies along the
main diagonal of the Weyl chamber, tracing the \emph{Heisenberg family} from the
identity ($\alpha = 0$, corner $O$) to the \textsc{swap} point ($\alpha = \pi/2$).
The maximally entangling point $\alpha = \pi/4$ sits at Weyl coordinates
$(\pi/4, \pi/4, \pi/4)$~---~the interior midpoint of this diagonal~---~and is
not locally equivalent to any of the standard fixed gates.

This trajectory is significant: fixed standard gates occupy single
\emph{points} in the Weyl chamber~---~\textsc{cnot} at $(\pi/2, 0, 0)$,
\textsc{iswap} at $(\pi/2, \pi/2, 0)$, \textsc{cz} at $(\pi/2, 0, 0)$
(locally equivalent to \textsc{cnot}).
The gate $\URA(\alpha)$ instead traces a \emph{continuous path} parameterized
by~$\alpha$, covering a range of entanglement capabilities rather than a
single fixed point.

\subsection{Entanglement Capability}
\label{subsec:entcap}

The \emph{entanglement capability} $\mathcal{E}(\alpha)$ of a two-qubit gate
is the maximum concurrence achievable over all product input
states~\cite{Zhang2003WeylChamber}:
\begin{equation}
  \mathcal{E}(\alpha) = \max_{\ket{\phi_1}\otimes\ket{\phi_2}}
    \mathcal{C}\!\left(\URA(\alpha)\ket{\phi_1\phi_2}\right).
  \label{eq:entcap}
\end{equation}
For $\URA(\alpha)$ with Weyl coordinates $(\alpha, \alpha, \alpha)$,
the entanglement capability is:
\begin{equation}
  \mathcal{E}(\alpha) = |\sin(2\alpha)|,
  \label{eq:entcap-result}
\end{equation}
which increases monotonically from $\mathcal{E}(0) = 0$ to
$\mathcal{E}(\pi/4) = 1$ and returns to $\mathcal{E}(\pi/2) = 0$
(the \textsc{swap} gate, which permutes states without generating entanglement).

The practical implication is direct: by setting $\alpha$ to the current
$\bar\alpha$ of the device's noise environment, the gate can be operated at
the entanglement capability precisely matched to the coherence available in
hardware~---~avoiding the overreach of fixed maximally entangling gates in
high-noise regimes.
Section~\ref{sec:sim} quantifies this advantage numerically.
