% ── APPENDIX A: FULL RCF PROOFS ──────────────────────────────────────────────

\section{Full Proofs for the Relational Coherence Framework}
\label{app:rct-proof}

\subsection{Mathematical Preliminaries}

Let $\mathcal{H}_S$ and $\mathcal{H}_E$ be finite-dimensional Hilbert spaces
with $\dim\mathcal{H}_S = d_S \geq 2$ and $\dim\mathcal{H}_E = d_E \geq d_S$.
The composite system occupies $\mathcal{H} = \mathcal{H}_S \otimes \mathcal{H}_E$.
Observables are self-adjoint elements of the C*-algebra
$\mathcal{B}(\mathcal{H})$ of bounded linear operators on $\mathcal{H}$.
States are density operators: positive trace-class operators
$\rho \in \mathcal{B}(\mathcal{H})$ with $\mathrm{Tr}[\rho] = 1$.

Fix orthonormal bases $\{\ket{s_i}\}_{i=1}^{d_S}$ and
$\{\ket{e_k}\}_{k=1}^{d_E}$ for $\mathcal{H}_S$ and $\mathcal{H}_E$
respectively.
Any pure state of $SE$ admits the Schmidt decomposition
\begin{equation}
  \ket{\psi_{SE}} = \sum_{j=1}^{r} \sqrt{\lambda_j}\,\ket{u_j}\otimes\ket{v_j},
  \label{eq:schmidt}
\end{equation}
where $r \leq \min(d_S, d_E)$ is the Schmidt rank, $\lambda_j > 0$,
$\sum_j \lambda_j = 1$, and $\{\ket{u_j}\}$, $\{\ket{v_j}\}$ are
orthonormal sets in $\mathcal{H}_S$ and $\mathcal{H}_E$ respectively.
The state is a product state iff $r = 1$; it is maximally entangled iff
$r = \min(d_S, d_E)$ and all $\lambda_j = 1/r$.

\subsection{Proof of Proposition~\ref{prop:indefiniteness} (Property Indefiniteness)}

The proposition statement in Sec.~\ref{sec:prop1} is the
linear-algebraic fact that $\bar\alpha(S,E)>0$ implies a non-diagonal
$\rho_S$ in the pointer basis, and that proof appears in
Sec.~\ref{sec:prop1}.
The Bell / Kochen--Specker material below is the auxiliary
no-hidden-variables result that supplies the physical content of
property indefiniteness against \emph{local non-contextual} hidden
variables; it is not a proof of the proposition as stated, but the
foundational anchor that justifies reading a non-diagonal $\rho_S$ as
the absence of a definite outcome rather than as ignorance of one.

\textbf{Auxiliary claim (no local non-contextual hidden variables).}
There is no \emph{local, non-contextual} value assignment
$v: \mathcal{O}(\mathcal{H}_S) \to \mathbb{R}$
mapping each self-adjoint observable $\hat{A}$ on $\mathcal{H}_S$ to a
definite eigenvalue $v(\hat{A}) \in \mathrm{spec}(\hat{A})$
consistently with all quantum predictions.
(Nonlocal hidden-variable theories such as Bohmian
mechanics~\cite{Bohm1980} evade this conclusion by accepting explicit
nonlocality; the result below constrains only local non-contextual
assignments.)

\textbf{Proof.}
I treat the two cases $d_S = 2$ and $d_S \geq 3$ separately.

\textit{Case $d_S = 2$ (qubit).}
Suppose, for contradiction, that definite pre-existing values
$A, A', B, B' \in \{+1,-1\}$ exist for all qubit observables and
depend only on local measurement settings.
The algebraic identity $|AB - AB' + A'B + A'B'| \leq 2$ holds for any
four such values, and averaging over the hidden-variable distribution
preserves the bound: for any joint state of two qubits and measurement
directions $\mathbf{a}, \mathbf{a}', \mathbf{b}, \mathbf{b}'$, the
CHSH combination satisfies
\begin{equation}
  \bigl|\langle\hat{A}\hat{B}\rangle - \langle\hat{A}\hat{B}'\rangle
    + \langle\hat{A}'\hat{B}\rangle + \langle\hat{A}'\hat{B}'\rangle\bigr|
  \leq 2.
\end{equation}
For the maximally entangled state $\ket{\Phi^+}$ with directions chosen at
$45^\circ$ intervals, quantum mechanics predicts a violation up to the
Tsirelson bound $2\sqrt{2}$~\cite{Tsirelson1980}, confirmed
experimentally~\cite{Aspect1982,Zeilinger2022Nobel}.
This contradiction establishes that no pre-existing value assignment exists
for $d_S = 2$.

\textit{Case $d_S \geq 3$.}
The Kochen--Specker theorem~\cite{KochenSpecker1967} proves that for
$d_S \geq 3$, there is no non-contextual hidden variable assignment
consistent with the quantum predictions for all projection-valued
measures on $\mathcal{H}_S$.
(A non-contextual assignment would give the same value to a projector
$\hat{P}$ regardless of which complete set of commuting observables it
is measured alongside~---~precisely the assignment required by any
hidden variable theory seeking to recover quantum predictions.)
No such assignment exists.

In both cases, properties are not possessed independently of measurement
context, i.e., independently of relational interaction. $\square$

\subsection{Proof of Proposition~\ref{prop:determination} (Relational Determination)}

The dynamical proof~---~the Lindblad calculation showing
$\bar\alpha(S,E;t) = \bar\alpha(S,E;0)\,e^{-\gamma t}$~---~is given in
Sec.~\ref{sec:prop2}.
Here I record an operational corollary that connects the dynamical
decay to the relational reading of measurement outcomes.

\textbf{Operational corollary.}
A definite pointer-basis outcome for an observable $\hat{A}$ on $S$
arises relative to an environment $E_i$ if and only if (a)~$S$ has
coupled to $E_i$ via an interaction Hamiltonian $H_{SE_i}$ that drives
the joint state from Schmidt rank~$1$ to rank~$r > 1$, and
(b)~einselection has subsequently driven the local pointer-basis
coherence $\bar\alpha(S,E_i;t)$ to zero.
Equivalently, the trajectory of $\bar\alpha(S,E_i;t)$ for a relevant
relational interaction is non-monotone: a transient rise (from
coupling) followed by exponential decay (from einselection,
Proposition~\ref{prop:determination}).

\textbf{Sketch.}

($\Rightarrow$)
Suppose either (a)~or~(b) fails.
If~(a) fails, $S$ has not coupled to $E_i$, the joint state remains a
product $\ket{\psi_S}\otimes\ket{e_0(t)}$ (Schmidt rank~$1$), and no
definite pointer outcome relative to $E_i$ is established by the
present interaction.
(If $\ket{\psi_S}$ is itself a pointer-basis superposition,
$\bar\alpha(S,E_i;t)$ may be nonzero in the basis induced by
$H_{SE_i}$, but the absence of $S$-$E_i$ correlation means $E_i$ holds
no record of any outcome of $\hat{A}$.)
If~(b) fails, $\bar\alpha(S,E_i;t) > 0$ persists, $\rho_S$ retains
off-diagonal coherence in the pointer basis, and by
Proposition~\ref{prop:indefiniteness} no definite pointer-state value
is assigned to $\hat{A}$.

($\Leftarrow$)
Suppose~(a)~and~(b) hold.
The joint unitary $U(t) = e^{-iH_{SE_i}t/\hbar}$ takes
$\ket{\psi_S}\otimes\ket{e_0}$ to a Schmidt rank-$r>1$ state, growing
$\bar\alpha(S,E_i;t)$ above zero in the einselected pointer basis.
Einselection~\cite{Zurek2003} then drives
$\bar\alpha(S,E_i;t)\to 0$ by Proposition~\ref{prop:determination},
leaving $\rho_S$ pointer-basis-diagonal: relative to $E_i$, $S$ has a
definite outcome. $\square$

\subsection{Proof of Proposition~\ref{prop:correspondence} (Coherence--Affinity Correspondence)}

\textbf{Claim.} (i)~$\bar\alpha = 0$ iff $\rho_S$ is diagonal in the
pointer basis (complete decoherence);
(ii)~$\bar\alpha = 1$ iff $\rho_S$ is a pure equal-weight pointer-basis
superposition $\ket{\psi} = d_S^{-1/2}\sum_k e^{i\phi_k}\ket{\pi_k}$;
(iii)~$\bar\alpha(t)$ is continuous and monotonically non-increasing
under the dephasing dynamics of Proposition~\ref{prop:determination}.

\textbf{Remark on local vs.\ joint scope.}~Throughout, $\bar\alpha(S,E)$
is built from the off-diagonal entries of the \emph{reduced} state
$\rho_S = \mathrm{Tr}_E[\rho_{SE}]$ in the pointer basis selected by
einselection (Definition~\ref{def:alpha-tensor}), not from the joint
state $\rho_{SE}$ in the Schmidt basis.
$\bar\alpha$ is therefore a measure of \emph{local} pointer-basis
coherence, not of $S$-$E$ entanglement; the two are inversely related
along the decoherence trajectory of
Proposition~\ref{prop:conservation}, but they are distinct objects.
In particular, a Bell state $\ket{\Phi^+} = (\ket{00}+\ket{11})/\sqrt{2}$
yields $\rho_S = \mathbf{I}/2$, hence $\bar\alpha(S,E) = 0$ at maximal
entanglement; conversely, a product superposition
$\ket{+}_S\otimes\ket{0}_E$ yields $\bar\alpha = 1$ at zero
entanglement.

\textbf{Proof.}
This is the proof presented in
Sec.~\ref{sec:prop3}; I restate it here for completeness.

\textit{Range bound.}
Since $\rho_S\geq 0$ with $\mathrm{Tr}\,\rho_S = 1$, the
Cauchy--Schwarz inequality gives
$\lvert\rho_{ij}\rvert^2 \leq \rho_{ii}\rho_{jj}$.
Summing over $i\neq j$,
\begin{equation*}
  \lVert\alpha\rVert_F^2
    \leq \sum_{i\neq j}\rho_{ii}\rho_{jj}
    = \Bigl(\sum_k\rho_{kk}\Bigr)^{\!2} - \sum_k\rho_{kk}^2
    = 1 - \sum_k\rho_{kk}^2
    \leq \frac{d_S - 1}{d_S},
\end{equation*}
using $\sum_k\rho_{kk}^2 \geq 1/d_S$ (power-mean / Cauchy--Schwarz).
Multiplying by $d_S/(d_S-1)$ and taking the square root yields
$\bar\alpha\leq 1$; non-negativity is immediate.

\textit{Part (i).}
$\bar\alpha = 0\Leftrightarrow\lVert\alpha\rVert_F = 0
 \Leftrightarrow \rho_{ij}=0$ for all $i\neq j$
 $\Leftrightarrow\rho_S$ pointer-basis-diagonal.

\textit{Part (ii).}
$\bar\alpha = 1$ requires equality in both bounds.
The Cauchy--Schwarz equality
$\lvert\rho_{ij}\rvert^2 = \rho_{ii}\rho_{jj}$ for every off-diagonal
pair forces $\mathrm{rank}\,\rho_S \leq 1$ (a pure state), and
$\sum_k\rho_{kk}^2 = 1/d_S$ forces $\rho_{kk} = 1/d_S$ for all $k$;
together,
$\ket{\psi} = d_S^{-1/2}\sum_k e^{i\phi_k}\ket{\pi_k}$.

\textit{Part (iii).}
Each off-diagonal magnitude satisfies
$\lvert\rho_{ij}(t)\rvert = \lvert\rho_{ij}(0)\rvert e^{-\gamma t}$
(Proposition~\ref{prop:determination}), so $\bar\alpha(t)$ is
continuous and strictly decreasing whenever $\bar\alpha(0)>0$.
$\square$

\textbf{Two-qubit reference values.}
For a two-qubit pure state
$\ket{\psi_{SE}} = c_{00}\ket{00} + c_{01}\ket{01}
                  + c_{10}\ket{10} + c_{11}\ket{11}$,
the local reduced state on qubit~$S$ is
\begin{equation*}
  \rho_S = \begin{pmatrix}
    \lvert c_{00}\rvert^2 + \lvert c_{01}\rvert^2 &
    c_{00}\overline{c_{10}} + c_{01}\overline{c_{11}} \\[2pt]
    \overline{c_{00}}c_{10} + \overline{c_{01}}c_{11} &
    \lvert c_{10}\rvert^2 + \lvert c_{11}\rvert^2
  \end{pmatrix},
\end{equation*}
so $\bar\alpha(S,E) = 2\,\lvert c_{00}\overline{c_{10}}
                      + c_{01}\overline{c_{11}}\rvert$
in the computational pointer basis.
This is \emph{not} equal to the concurrence
$\mathcal{C} = 2\,\lvert c_{00}c_{11} - c_{01}c_{10}\rvert$ in general:
$\bar\alpha$ probes pointer-basis coherence in $\rho_S$, whereas
$\mathcal{C}$ probes the entanglement of $\rho_{SE}$. The Bell state
$\ket{\Phi^+}$ has $\bar\alpha = 0$ while $\mathcal{C} = 1$;
$\ket{+}_S\otimes\ket{0}_E$ has $\bar\alpha = 1$ while $\mathcal{C} = 0$.

\subsection{Proof of Proposition~\ref{prop:conservation} (Coherence Redistribution under Unitary $S$--$E$ Evolution)}

The full three-part proof is given in Sec.~\ref{sec:prop4}.
I record here only the elementary purity identity that anchors
part~(i), for completeness.

\textbf{Claim (part (i)).}
For a closed system with unitary evolution $U(t)$,
$\mathrm{Tr}[\rho_{SE}(t)^2] = \mathrm{Tr}[\rho_{SE}(0)^2]$ for all $t$.

\textbf{Proof.}
\begin{align*}
  \mathrm{Tr}\bigl[\rho_{SE}(t)^2\bigr]
    &= \mathrm{Tr}\bigl[U(t)\rho_{SE}(0)U^\dagger(t)\cdot
                        U(t)\rho_{SE}(0)U^\dagger(t)\bigr] \\
    &= \mathrm{Tr}\bigl[U(t)\rho_{SE}(0)^2 U^\dagger(t)\bigr]
      \qquad (\text{since } U^\dagger U = \mathbf{I}) \\
    &= \mathrm{Tr}\bigl[\rho_{SE}(0)^2\bigr]
      \qquad (\text{cyclicity of trace}).
\end{align*}
The apparent purity loss in the reduced state
$\rho_S = \mathrm{Tr}_E[\rho_{SE}]$ reflects transfer of coherence
into $S$-$E$ correlations~---~the joint state remains pure, and
$\mathrm{Tr}[\rho_{SE}^2] = 1$ throughout.
The ``lost'' coherence is not destroyed; it is encoded in the
relational network $\Gamma_S$ (Sec.~\ref{sec:rqm}) as $S$-$E$
entanglement. $\square$

\subsection{Weyl Chamber Coordinates for $\URA(\theta)$}

The local invariants of a two-qubit gate $U \in \mathrm{SU}(4)$ are
extracted from the matrix
$M(U) = m(U)^T m(U)$, where $m(U) = U_B^T\, U\, U_B$ is the gate
represented in the Bell basis~\cite{Zhang2003WeylChamber} via

$U_B = \frac{1}{\sqrt{2}}\begin{pmatrix}1&0&0&i\\0&i&1&0\\0&i&{-1}&0\\1&0&0&{-i}\end{pmatrix}$.

For $\URA(\theta)$ defined by Eq.~\eqref{eq:URA}, direct computation gives
$M = \mathrm{diag}(e^{-i\theta}, e^{i\theta}, e^{i\theta}, e^{-i\theta})$
in the Bell basis, yielding Weyl coordinates:
\begin{equation}
  (c_1, c_2, c_3) = (\theta, \theta, \theta), \quad \theta \in [0, \pi/2].
  \label{eq:weyl-full}
\end{equation}
This trajectory lies along the main diagonal $(c_1, c_2, c_3) = (\theta, \theta, \theta)$
of the Weyl chamber, the isotropic-Heisenberg-exchange family, confirming
that $\URA$ is an isotropic exchange gate continuously interpolating from
the identity ($\theta = 0$) to the \textsc{swap} gate ($\theta = \pi/2$),
with maximum entanglement capability at $\theta = \pi/4$ (the
$\sqrt{\textsc{swap}}$ point of the Weyl chamber, locally inequivalent to
\textsc{iswap}, which sits at $(\pi/2, \pi/2, 0)$).

\subsection{Simulation Parameters}
\label{app:sim-params}

All simulations use the following configuration unless otherwise noted.

\textbf{Qiskit (IBM):} \texttt{qiskit-terra} v0.45, \texttt{qiskit-aer} v0.13.
Two-qubit system; $10^4$ shots per circuit; state vector simulation for
process fidelity; density matrix simulation for noise models.
For the noise-model benchmarks reported in
Tables~\ref{tab:fidelity}--\ref{tab:fidelity-all}, $\URA(\theta)$ is
applied as a single two-qubit unitary with gate-time-proportional
effective error rate $p_{\rm eff}(\theta)$ per Eq.~\eqref{eq:peff};
the closed-form depolarizing values match the simulator output to
numerical precision under this single-unitary model.
On real hardware without direct exchange calibration, transpilation to
a fixed native gate set such as $\{R_Z, \sqrt{X}, \textsc{cx}\}$ would
introduce decomposition overhead that can attenuate or eliminate the
fidelity advantage (cf.\ Sec.~\ref{sec:disc}).

\textbf{Cirq (Google):} \texttt{cirq-core} v1.3.
Same qubit configuration; \texttt{DensityMatrixSimulator} with noise models
applied as \texttt{ConstantQubitNoiseModel}.

\textbf{Noise models}
(per-full-duration two-qubit gate; effective rate scaled by gate-time
factor $\theta/(\pi/4)$ for $\URA(\theta)$ per Eq.~\eqref{eq:peff}):
\begin{enumerate}
  \item \emph{Depolarizing}: $p_{\rm base} \in \{0.01, 0.02, 0.05, 0.10\}$.
  \item \emph{Dephasing (phase damping)}: rate $\gamma \in \{0.02, 0.05, 0.10, 0.20\}$.
  \item \emph{Amplitude damping}: $T_1$ relaxation probability
        $p_{\rm AD} \in \{0.02, 0.05, 0.10, 0.20\}$.
\end{enumerate}

\textbf{Fidelity definitions.}
The process fidelity $F_{\rm proc}(W,V) = \mathrm{Tr}[\chi_V^\dagger\,\chi_W]$
is computed from the Choi--Jamio\l{}kowski matrices $\chi_V$
(ideal) and $\chi_W$ (noisy implementation).
The \emph{average gate fidelity} reported in Tables~\ref{tab:fidelity}
and~\ref{tab:fidelity-all} is the standard Haar average over input
states,
$F_{\rm avg}(W,V) = (d\,F_{\rm proc}(W,V) + 1)/(d+1)$ with $d = 4$,
which for the global depolarizing channel
$W(\rho) = (1-p_{\rm eff})V\rho V^\dagger + p_{\rm eff}\,\mathbf{I}_4/4$
reduces to $F_{\rm avg} = 1 - 3\,p_{\rm eff}/4$.

\textbf{Baseline gates:} \textsc{cnot} = $\mathrm{CX}$ gate at $(c_1,c_2,c_3) = (\pi/2,0,0)$;
\textsc{iswap} at $(\pi/2,\pi/2,0)$; \textsc{cz} at $(\pi/2,0,0)$ (locally
equivalent to \textsc{cnot}).
All baselines implemented via native Qiskit/Cirq gate objects.
